Research Experience

### 2010-04-25 doc/thesis/thesis.tex
more work on the lit review, and some cleanup

\documentclass{article}
\usepackage[left=1cm,top=2cm,right=1cm,nohead,nofoot]{geometry}

%\usepackage{times}
%\usepackage{tikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{cancel}

%\doublespacing

@@

\begin{document}
\section{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity for the majority of the brain.
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite its prevalence, there have been relatively few developments
in the actual analysis of FMRI images. A steady stream of studies have built
on the original BOLD signal derivation first described in \cite{Ogawa}, 
from the Baloon model first proposed by \cite{Buxton1998}
all the way to full fully autonomous system of equations \cite{Riera2004}. And while
there have been numerous forks in the model, enough in fact to make an entire paper
studying the differences,\cite{Deneux2006}, it is widely known that all these
models have quantitatively less bias error than General Linear Model which is
typically employed today. Then again, depending on the model there may be between
seven \cite{Riera2004} and fifteen \cite{Zheng2002} parameters per voxel to
be optimized. Clearly there is a significant risk of error due to variance
with so many degrees of freedom, not to mention a significantly increased
computation cost. In this thesis I demonstrate the use of a 
particle filter as a means of addressing these problems.

@@

\subsection{BOLD Physiology}
It is well known that the two types of hemoglobin act as a contrast agents in 
EPI imaging
\cite{Buxton1999}, \cite{WEISSKOFF1994}, \cite{Ogawa}, however the connection
between DeHb/O2Hb and neural activity is non-trivial. Intuitively, increased 
metabolism will increase Deoxyhemoglobin, however blood vessels are quick
to compensate by increasing local blood flow. Increased inflow will of course
preceed increased outflow, and increased inflow is encouraged by loosening 
capilary beds. Both of these factors drive increased storage capacity.
Since the local MR signal depends on the ratio of Deoxyhemoglobin to Oxygenated
Hemoglobin, increased volume of blood can certainly effect this ratio if 
metabolism doesn't exactly match the increased inflow of oxygenated blood.
This was the impetus
for the ground breaking balloon model (\cite{Buxton1999}) and windkessel
model (\cite{Mandeville1999}). These models derive from first principals
the increased deoxyhemoglobin ratio and volume of capillaries based on flow.
These were the first two attempts to quantitatively account for the shape of the 
BOLD signal as a consequence of the lag between the cerebral blood flow (CBF) 
and the cerebral blood volume (CBV). In fact \cite{Buxton1999} went to far as
to show that a simple, well chosen blood flow waveform coupled with a simple
cerebral metabolic rate of oxygen (CMRO2) curve, in the context of a balloon 
model, could fully account for the BOLD signal.

Although complex
models for metabolism exist (\cite{Zheng2002}), most recent studies have
focused on their ability to explain the prolonged BOLD post stimulus 
undershoot \cite{Donahue2009}, \cite{Yacoub2006}. This is because \cite{Buxton2004}
and later \cite{Riera2004} showed that the main portion of the signal may be 
accurately estimated by a simple blood flow locked expression of the CMRO2. 
Although \cite{Deneux2006} did not deal extensively with prolonged post
stimulus undershoot, the comparisons made in that publication showed minimal
improvement from separate expressions of CMRO2. For some time there was 
debate regarding the cause for the post-stimulus undershoot often seen in 
FMRI studies. The problem stemmed from the fact that there were many studies
showing CBV and CBF returning to baseline well before the BOLD signal. In fact
this result seemed to be confirmed by several studes: \cite{Frahm2008}, \cite{Donahue2008}, 
\cite{Lu2004}.

@@

In fact, the original Buxton
paper demonstrated how minimal changes in the flow expression could
result in vastly different recovery times of the baseline BOLD signal, and
in \cite{Mandeville1999} a cause for delayed CBF recovery was proposed, based
on hysterous in the compliance of capillaries. \cite{Behzadi2005} provided
additional incite into this possibility, going so far as proposing a simple
yet complete model and explanation for delayed compliance of neural capillaries.
Those simple changes can result in error levels on par with the more 
complex CMRO2 inclusive models, which was one of the major findings of \cite{Deneux2006}. 
This was reiterated in \cite{Chan2009}, which concluded that many of
the studies claiming a CBV recovery well before the BOLD signal 
(\cite{Frahm1996}, \cite{Kruger1996}, \cite{Lu2004}) 
could actually have been detecting arterial blood flow, which does not significantly effect the
BOLD signal. 
In the end, it may be impossible to refute the
possibility of a prolonged period of increased metabolic activity after activation,
however, much of the BOLD post stimulus undershoot may be explained by a 
prolonged CBV undershoot, as shown in \cite{Chan2009}. Because of the greatly
increased complexity of CMRO2 modeling, compared with the relative simplicity
of the Behzadi model, in most cases it is not worth using the more advanced CMRO2
model unless the metabolic rate is an end goal.

The physiology that leads to the BOLD signal has been well studied over the 
past decade and a half, but there is still some question about the origins
of certain signal features. In the seminal paper \cite{Buxton1999}, the original
Balloon model was proposed. Buxton et. al showed that all of the features
of the BOLD signal may be explained by lag between ... . While some uncertainty remains (for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>) generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events that locally 
alters the composition of the blood. Inactive neurons can be thought of
as a slingshot cocked and ready to fire. As soon as a signal moves up
the axon and causes the neuron to fire (and thus changes the state
of a membrane at that location), ions quickly move across the altered
membrane to compensate for a high charge imbalance between the interior
and exterior of the cell. This process, similar to allowing a stretched
rubber band to contract leaves the system at a lower energy state. This cannot
last though, because the neuron needs to be ready to fire again rather quickly.
To return the cell to a ready position the membrane becomes impermeable
to the ions again, and then begins pumping ions back into the cell. This
process takes a large amount of energy, whereas the actual firing takes
very little energy. Thus, after firing, glucose is burned, removing oxygen
from the blood and thus causing a dip in the ratio of oxygenated hemoglobin to 
de-oxygenated hemoglobin. This is the first potentially measurable effect
that MRI can see, although it usually lasts fewer than 2 seconds making it
difficult for FMRI to catch. As a result of the decreased amount of local oxygen,
the capillaries compensate by increasing blood flow to that region. Because,
of the quickly increased blood flow into local capillaries, the blood
volume increases in addition to the local oxygen content. This leads to a Windkessel
effect which further lags the normalization of oxygen content. The effect of
all this is an overshoot in the oxygen content above the initial level. After
the work of recharging is done, there may be a prolonged undershoot lasting
as much as 90 seconds \cite{spatial_undershoot}, though the reason for this
is debated \cite{origin_poststim}.

A large number of models have been proposed for the BOLD signal with varying
amounts of complexity. The simplest model is the so called "Balloon Model"
which first proposed the windkessel effect as a factor the BOLD response.
The model we will use is the model proposed by Buxton Et. Al. and later used
in Riera et. al. \cite{Riera}. The model has four state variables, $s$, $f$,
$v$, $q$, representing flow inducing signal, cerebral blood flow, cerebral
blood volume and deoxyhemaglobin to hemaglobin ratio, respectively. The 
state variables change over time, given by the state evolution equations:
\begin{equation}
\dot{s} = \epsilon u -  
\end{equation}

@@

\bibliographystyle{abbrvnat}
\bibliography{references}

>>>>

\documentclass{article}
\usepackage[left=3cm,top=2cm,right=3cm,nohead,nofoot]{geometry}

\usepackage{times}
%\usepackage{tikz}
%\usetikzlibrary{trees}
%\usetikzlibrary{shapes,snakes}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{cancel}

\doublespacing

@@

\begin{document}
\section{Introduction}
Functional Magnetic Resonance Imaging (FMRI) is a powerful tool in the analysis
of neural activity. Despite its rather limited temporal resolution, FMRI is still
the best way of measuring neural activity for the majority of the brain.
Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is relatively quick and cheap, and its analysis straight forward.
Because of these benefits, FMRI continues to be crucial to the study of human 
cognition. Despite its prevalence, there have been relatively few developments
in the actual analysis of FMRI images. A steady stream of studies have built
on the original BOLD signal derivation first described in \cite{Ogawa}, 
from the Baloon model first proposed by \cite{Buxton1998}
all the way to full fully autonomous system of equations \cite{Riera2004}. And while
there have been numerous forks in the model, enough in fact to make an entire paper
studying the differences,\cite{Deneux2006}, it is widely known that all these
models have quantitatively less bias error than General Linear Model which is
typically employed today. Then again, depending on the model there may be between
seven \cite{Riera2004} and fifteen \cite{Zheng2005} parameters per voxel to
be optimized. Clearly there is a significant risk of error due to variance
with so many degrees of freedom, not to mention a significantly increased
computation cost. In this thesis I demonstrate the use of a 
particle filter as a means of addressing these problems.

@@

\subsection{BOLD Physiology}
It is well known that the two types of hemoglobin act as a contrast agents in 
EPI imaging
\cite{Buxton1998}, \cite{WEISSKOFF1994}, \cite{Ogawa}, however the connection
between Deoxyhemoglobin/Oxygenated Hemoglobin and neural activity is non-trivial. 
Intuitively, increased 
metabolism will increase Deoxyhemoglobin, however blood vessels are quick
to compensate by increasing local blood flow. Increased inflow will of course
preceed increased outflow, and increased inflow is accomplished by loosening 
capilary beds. Both of these factors drive increased storage capacity.
Since the local MR signal depends on the ratio of Deoxyhemoglobin to Oxygenated
Hemoglobin, increased volume of blood can certainly effect this ratio if 
metabolism doesn't exactly match the increased inflow of oxygenated blood.
This was the impetus
for the ground breaking balloon model (\cite{Buxton1998}) and windkessel
model (\cite{Mandeville1999}). These models derive from first principals
the increased deoxyhemoglobin ratio and volume of capillaries based on a given flow.
These were the first two attempts to quantitatively account for the shape of the 
BOLD signal as a consequence of the lag between the cerebral blood volume (CBV) 
and the cerebral blood flow (CBF). In fact \cite{Buxton1998} went to far as
to show that a simple, well chosen blood flow waveform coupled with a square 
wave cerebral metabolic rate of oxygen (CMRO2) curve, in the context of a balloon 
model, could fully account for the BOLD signal.

Although \cite{Buxton1998} showed that a well chosen flow waveform could 
explain much of the BOLD signal, there was still a matter of proposing a
realistic waveform for the CBF and for the CMRO2. \cite{friston2000} gave
a reasonable and simple
expressoin for CBF input,$f$, based on a flow inducing signal, $s$, 
\begin{eqnarray}
\dot{s} &=& \epsilon u(t) - \frac{s}{\tau_s} - \frac{f - 1}{\tau_f} \\
\dot{f} &=& s
\end{eqnarray}
where $\epsilon$ is a neuronal efficiency term, $u(t)$ is a stimulus, and $\tau_f$, $\tau_s$
are both time constants. In \cite{Buxton2004} the final piece of the simple balloon
model was put into place, by describing the CMRO2 as a constant multiple of
the CBF (the inflow of blood). This completed the basic balloon model, and 
was well summarized in \cite{Riera2004}. 
\begin{eqnarray}
\dot{v} &=& \frac{1}{\tau_0}(f - v^\alpha)\\
\dot{q} &=& \frac{1}{\tau_0}(\frac{f(1-(1-E_0)^f)}{E_0} - \frac{q}{v^{1-1/\alpha}})
\end{eqnarray}
where $v$ is normalized cerebral blood volume (CBV), and $q$ is the normalized
local deoxyhemoglobin/oxygenated hemoglobin ratio, $E_0$ is the resting metabolic
rate and $\alpha$ is Grubb's parameter controling the balloon model. 
\cite{Obata2004} refined the readout equation of the BOLD signal based on the
deoxyhemoglobin content (q) and local blood volume (v), resulting in the
final BOLD equation:
\begin{eqnarray}
y   &=& V_0((k_1 + k_2)(1-q) - (k_2 + k_3)(1-v))\\
k_1 &=& 4.3 \times \nu_0 \times E_0 \times TE = 2.8\\
K_2 &=& \epsilon_0 \times r_0 \times E_0 \times TE = .57\\
k_3 &=& \epsilon_0 - 1 = .43
\end{eqnarray}
Where $\nu_0 = 40.3 s^{-1}$  is the frequency offset in Hz for fully
deoxygenated blood (at 1.5T), $r_0 = 25 s^{-1}$  is the slope relating
change in relaxation rate with change in blood oxygenation, and
$\epsilon_0 = 1.43$ is the 
ratio of signal MR from intravascular to extravascular at rest. Although,
obviously these constants change with experiment ($TE$, $\nu_0$, $r_0$),
patient, and brain 
region ($E_0$, $r_0$), often the estimated values taken from \cite{Obata2004} are used
as constant ($k_1 + k_2 = 3.4$, and $k_2+k_3 = 1$).
While this model is in a sense complete, it is far from perfect. The major
problem often brought up with this version of the BOLD model is that it
does not represent the so called "post-stimulus undershoot" well.

Although complex
models for metabolism exist (\cite{Zheng2005}), most recent studies have
focused on their ability to explain the prolonged BOLD post stimulus 
undershoot \cite{Donahue2009}, \cite{Yacoub2006}. This is because \cite{Buxton2004}
and later \cite{Riera2004} showed that the main portion of the signal may be 
accurately estimated by a simple blood flow locked expression of the CMRO2.. 
Although \cite{Deneux2006} did not deal extensively with prolonged post
stimulus undershoot, the comparisons made in that publication showed minimal
improvement from separate expressions of CMRO2, in comparison to the much 
increased complexity. \cite{Deneux2006} did show that by simply adding viscoelastic
terms, first proposed in \cite{Buxton2004}, that a slowed return to baseline for
the BOLD signal is possible to model. However, viscoelastic effects primarily
control CBV, which some studies have claimed returns to baseline \emph{before}
the BOLD signal. If this is the case, then CBV cannot be the cause of the 
post-stimulus undershoot. Even now there is some debate about the cause of 
the post-stimulus undershoot. The BOLD post-stimulus undershoot is common,
and typically lasts longer for longer activations (\cite{Chen2009}), and in
fact varies across regions of the brain (\cite{Mandeville2001}, \cite{Yacoub2006})
debate regarding the cause for the post-stimulus undershoot often seen in 
FMRI studies. The problem stemmed from the fact that there were many studies
showing CBV and CBF returning to baseline well before the BOLD signal. In fact
this result seemed to be confirmed by several studes: \cite{Frahm2008}, \cite{Donahue2009}, 
\cite{Lu2004}.

@@

In fact, the original Buxton
paper demonstrated how minimal changes in the flow expression could
result in vastly different recovery times of the baseline BOLD signal, and
in \cite{Mandeville1999} a cause for delayed CBF recovery was proposed, based
on hysterous in the compliance of capillaries. \cite{Behzadi2005} provided
additional incite into this possibility, going so far as proposing a simple
yet complete model and explanation for delayed compliance of neural capillaries.
Those simple changes can result in error levels on par with the more 
complex CMRO2 inclusive models, which was one of the major findings of \cite{Deneux2006}. 
This was reiterated in \cite{Chen2009}, which concluded that many of
the studies claiming a CBV recovery well before the BOLD signal 
(\cite{Frahm1996}, \cite{Kruger1998}, \cite{Lu2004}) 
could actually have been detecting arterial blood flow, which does not significantly effect the
BOLD signal. 
In the end, it may be impossible to refute the
possibility of a prolonged period of increased metabolic activity after activation,
however, much of the BOLD post stimulus undershoot may be explained by a 
prolonged CBV undershoot, as shown in \cite{Chen2009}. Because of the greatly
increased complexity of CMRO2 modeling, compared with the relative simplicity
of the Behzadi model, in most cases it is not worth using the more advanced CMRO2
model unless the metabolic rate is an end goal.

The physiology that leads to the BOLD signal has been well studied over the 
past decade and a half, but there is still some question about the origins
of certain signal features. In the seminal paper \cite{Buxton1998}, the original
Balloon model was proposed. Buxton et. al showed that all of the features
of the BOLD signal may be explained by lag between ... . While some uncertainty remains (for instance in
relation to the BOLD post-stimulus undershoot <papers relating to back and
forth on post-stimulus undershoot>) generally the model fits reality very 
well. As with every piece of tissue in the body, the brain requires oxygen
to extract energy from glucose in the blood. This process of removing oxygen
from red blood cells sets into motion a chain of events that locally 
alters the composition of the blood. Inactive neurons can be thought of
as a slingshot cocked and ready to fire. As soon as a signal moves up
the axon and causes the neuron to fire (and thus changes the state
of a membrane at that location), ions quickly move across the altered
membrane to compensate for a high charge imbalance between the interior
and exterior of the cell. This process, similar to allowing a stretched
rubber band to contract leaves the system at a lower energy state. This cannot
last though, because the neuron needs to be ready to fire again rather quickly.
To return the cell to a ready position the membrane becomes impermeable
to the ions again, and then begins pumping ions back into the cell. This
process takes a large amount of energy, whereas the actual firing takes
very little energy. Thus, after firing, glucose is burned, removing oxygen
from the blood and thus causing a dip in the ratio of oxygenated hemoglobin to 
de-oxygenated hemoglobin. This is the first potentially measurable effect
that MRI can see, although it usually lasts fewer than 2 seconds making it
difficult for FMRI to catch. As a result of the decreased amount of local oxygen,
the capillaries compensate by increasing blood flow to that region. Because,
of the quickly increased blood flow into local capillaries, the blood
volume increases in addition to the local oxygen content. This leads to a Windkessel
effect which further lags the normalization of oxygen content. The effect of
all this is an overshoot in the oxygen content above the initial level. After
the work of recharging is done, there may be a prolonged undershoot lasting
as much as 90 seconds \cite{Yacoub2006}, though the reason for this
is debated \cite{Chen2009}.

A large number of models have been proposed for the BOLD signal with varying
amounts of complexity. The simplest model is the so called "Balloon Model"
which first proposed the windkessel effect as a factor the BOLD response.
The model we will use is the model proposed by Buxton Et. Al. and later used
in Riera et. al. \cite{Riera2004}. The model has four state variables, $s$, $f$,
$v$, $q$, representing flow inducing signal, cerebral blood flow, cerebral
blood volume and deoxyhemaglobin to hemaglobin ratio, respectively. The 
state variables change over time, given by the state evolution equations:
\begin{equation}
\dot{s} = \epsilon u -  
\end{equation}

@@

\bibliographystyle{apalike}
%\bibliographystyle{abbrvnat}
%\bibliographystyle{abbrv}
\bibliography{library}
